# Full Code Documentation
The full code was not able to be rendered in this website. This pdf provides the full documentation and codes used, while images and references are included on the main webpage.

## Methodology
To test our hypotheses, the locations of fairy barrows will be compared to land cover data from 1990 to determine whether fairies prefer a certain habitat (first-order clustering). To test whether fairies cluster near stone circles (second-order clustering), rather than a particular habitat, we will compare the average distance of fairy barrows to nearby stone circles. The preferred habitat for stone circle construction will also be tested with the same land cover data, as we have very little information about fairy habits. If any preferred habitats are found, land cover data from 2018 will be compared to the 1990 dataset to produce a change matrix, and identify whether fairy habitat in the Repulic of Ireland remains intact. 

# Study Environment Preparation

This study was conducted using R, with the packages tidyverse, sf, mapview, spatstat, stars, rnaturalearth, terra, tidyterra, and pbapply.

The Coordinate Reference System (CRS) for this study is the Irish Transverse Mercator (ITM) projection, with EPSG code 2157. All data is reprojected or transformed into this CRS for calculations and visualizations. (cite)

```{r}
#| label: itinialiigjgjg
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
library(pbapply) |> suppressPackageStartupMessages()
set.seed(6805)
```

# Fairy Point Data Preparation
Barrows and Stone Circles data were collected from the Irish Archaeology website, downloaded in kml format from the associated Interactive Google Map, and converted to CSV. The CSV was also cleaned in Microsoft Excel to retain all downloaded information in a usable format for the sf library.  Metadata for these datasets can be found in the Appendix. 

```{r}
#| label: barrows data prep

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

```{r}
#| label: circles data prep

stone_circles_df <- read_csv('data/raw/stone-circles.csv')
stone_circles_df <- stone_circles_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "stone circle")

stone_circles_sf <- stone_circles_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(stone_circles_sf,100)) #showing the first 100 points
head(stone_circles_sf)

```

```{r}
#| label: combining them for visualization
fairies_sf <- rbind(barrows_sf, stone_circles_sf)
fairies_sf <- fairies_sf |> sf::st_transform(2157)

fairies_map <- mapview(fairies_sf, zcol="type")

```

The observation window for this study is the Natural Earth Data Republic of Ireland sf object at 1:10m resolution. All calculations are constrained to the bounds of this polygon.

```{r}
#| label: ireland observation window

ireland <- ne_countries(country = "ireland", scale="large") |> select(name,geometry)
ireland <- ireland |> sf::st_transform(2157)
mapview(ireland) + fairies_map

```

Each of the point layers were converted into a spatstat ppp (planar point pattern) object and plotted inside the Ireland observation window for the remainder of the calculations.

```{r}
#| label: barrows ppp 1
barrows_sf <- barrows_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
barrows_sf <- barrows_sf[st_within(barrows_sf, ireland, sparse = FALSE), ] 
barrows_ppp <- as.ppp(barrows_sf, W=as.owin(ireland)) 
barrows_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

```{r}
#| label: stone circles ppp 1
stone_circles_sf <- stone_circles_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
stone_circles_sf <- stone_circles_sf[st_within(stone_circles_sf, ireland, sparse = FALSE), ] 
stone_circles_ppp <- as.ppp(stone_circles_sf, W=as.owin(ireland)) 
stone_circles_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

# Land Cover Data Preparation

Land cover data from 1990 and 2018 was collected from the EU Copernicus component of the European Union's space program. It is produced by the Coordination of Information on the Environment (CORINE) program, which provides a uniform, continent-wide land cover inventory with 44 thematic classes called the CORINE Land Cover product. Full metadata of the 44 classes and their associated reclassification can be found in the Appendix.

```{r}
#| label: reading in the rasters againnnn

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)



# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)



```

```{r}
#| label: reclassification 1990 2 at the beginning
rclmat <- matrix(c(1, 11, 1,
                   11.999, 22, 2,
                   22.999, 25, 3,
                   25.999, 27, 4,
                   27.999, 29, 3,
                   29.999, 34, 5,
                   34.999, 39, 6,
                   39.999, 44, 7
                   ), ncol=3, byrow=TRUE) #decimals so that it catches all possible values

rc_lc_1990 <- classify(lc_1990, rclmat, include.lowest=TRUE)

```

```{r}
#| label: visualizing the lc raster 1990 at teh beginning (make pretty later)
#creating the color library for the classes
lc_colors <- data.frame(
  value = c(1, 2, 3, 4, 5, 6, 7),
  color = c("#8B0000", "#F5DEB3", "#006400", "#556B2F","#8B7355", "#7FFFD4", "#1E90FF")
)

lc_labels <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")

plot(lc_1990, type = "classes", main = "Land Cover: 1990")
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")

```

Full land cover reclassification table available in the Appendix(link?). Original land cover had 44 discrete classes, which was reduced to the following 7 discrete classes.

| New Land Cover Category | New Cell Value |
|----------|------|
| Dense Human Habitat | 1 |
| Human Agriculture | 2 |
| Forests | 3 |
| Grass and Shrublands | 4 |
| Rocky, Barren, or Snowy | 5 |
| Wetlands | 6 |
| Water Bodies | 7 |

Our second hypothesis will be average distance from fairy circles AND what habitats those fairy circles were built in (using a similar simmulation probably)

Turning the classified raster into polygons so we can start doing the simulations

```{r}
#| label: raster into polygons
lc_poly_1990 <- as.polygons(rc_lc_1990)

# turning it into an sf object
lc_poly_sf <- st_as_sf(lc_poly_1990)

#dissolving the polygons so we can do an area calculation
lc_poly_sf <- lc_poly_sf |>
  group_by(U2000_CLC1990_V2020_20u1) |>
  summarise()

lc_poly_sf <- st_transform(lc_poly_sf, 2157)
lc_poly_sf <- st_intersection(lc_poly_sf, ireland) #clipping the polygons so the borders are the same as my viewing window

mapview(lc_poly_sf, zcol = "U2000_CLC1990_V2020_20u1")

```

## Hypothesis Testing
The following section has the details for each hypothesis test. We will test the following hypotheses:
Null Hypothesis: Fairies in the Republic of Ireland display no significant clustering in home locations, indicating no preference in habitat.

Hypothesis 1: Fairies in the Republic of Ireland prefer forested habitats.

Hypothesis 2: Fairies in the Republic of Ireland prefer to live near stone circles.

Hypothesis 3: Stone circles were built in a certain habitat, reflecting the nature of the fairy realm on the other side.

# Hypothesis 1: Barrow Preferred Habitats
To test this hypothesis, we will generate a habitat 'intensity' funcion and use it to determine whether fairies cluster in any of the major land cover types: Dense Human Habitat, Human Agriculture, Forests, Grass and Shrublands, Rocky, Barren, or Snowy, Wetlands, or Water Bodies. The number of barrows found in each habitat type will form the initial observations, which we will compare to a Monte Carlo simulation. The Monte Carlo simulation will generate 999 similar point patterns of hypothetical barrows, proportionally generated into each habitat type. We will then compare the probability of a barrow appearing in each habitat type to the true proportion of real barrows we have observed. 

Below, we see the null distribution of the likely number of barrows in forest habitats, shown in grey bars, and the true number of barrows found in forests in the existing data, shown by a green line. If we had the 'expected' number of barrows in forest habitats for the amount of forested area in Ireland, the green line would be somewhere within the grey null distribution. However, this chart shows that there are far *less* barrows in forested areas, indicating that fairies actually *avoid* building their barrow homes in forests. 

```{r}
#| label: barrow counts by habitat
#making lc into a tesselation
lc_list <- split(lc_poly_sf, lc_poly_sf$U2000_CLC1990_V2020_20u1) #grabbing individual polygons
lc_tiles <- lapply(lc_list, as.owin) #each one as a window

#tesselation of the polygon windows
lc_tess <- tess(tiles = lc_tiles)

#now the test statistic
obs_barrow_counts <- quadratcount(barrows_ppp, tess=lc_tess) |> as.vector()
names(obs_barrow_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
obs_barrow_counts <- obs_barrow_counts |> as_tibble_row()
```

The test statistic: Referencing the literature, it is believed that fairies prefer forested habitats, which offer readily available hollow trees, babbling streams, and mossy grottos. Our first hypothesis, therefore, is that fairies prefer to make their homes in forests.

```{r}
obs_barrow_counts
```


habitat 'intensity', which gives the proportional probability of any point existing in each habitat type (based on area)

```{r}
#new habitat prob intensity, to try and make this run faster
tile_areas <- sapply(lc_tiles, area.owin) #the area of each tile
tile_probs <- tile_areas / sum(tile_areas) #getting the proportions

habitat_prob_intensity <- function(x, y) {
  coords <- cbind(x, y)
  cell_vals <- terra::extract(rc_lc_1990, coords)[,1] #pulls out the habitat value the point lands on top of
  cell_vals[is.na(cell_vals) | is.nan(cell_vals) | cell_vals < 1 | cell_vals > length(tile_probs)] <- 1 #prevent error by keeping everything in range

  probs <- tile_probs[cell_vals] #assigning the probabilties
  
  probs[is.na(probs)] <- 0 #preventing errors with NA by making them 0
  return(probs)
}
```

## Monte Carlo Simulations: Barrows

```{r}
#testing and developing the final function

num_barrows <- nrow(barrows_sf)#to set number of points to sample

#creating a function that will plot barrows around ireland according to the area proportions of the various habitats
#and checking the runtime to make sure we can do the full 999 sims
runtime <- system.time({gen_barrows_sim <- function() {
  barrows_sim <- rpoint(n=num_barrows,f=habitat_prob_intensity, win=barrows_ppp$window)
  return(barrows_sim) 
}
single_barrows_sim_ppp <- gen_barrows_sim()
#IT WORKED!!! YAYYYYYY
single_barrow_sim_counts <- quadratcount(single_barrows_sim_ppp, tess=lc_tess) |> as.vector()
names(single_barrow_sim_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
single_barrow_sim_counts <- single_barrow_sim_counts |> as_tibble_row()})
single_barrow_sim_counts
runtime
```



This is what I ran to do the monte carlo simulations. I never want to run that again, but here is the code
OK and I do have to run this again and save sims_list so that we have the points for the distance calculation, unfortunately

library('pbapply')
#Now we are running 999 simulations wahoooo

compute_quadrat_counts <- function(sim_ppp) { #making a function to apply to all the sims
  sim_counts <- quadratcount(sim_ppp, tess=lc_tess) |> as.vector()
  names(sim_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
  return(sim_counts)
}

gen_sims_ppp <- function(num_sims) { #function to generate 999 ppp simulations
  barrows_sims <- spatstat.random::rpoint(
    n = num_barrows,
    f = habitat_prob_intensity,
    win = barrows_ppp$window,
    nsim = num_sims
  )
  return(barrows_sims)
}

sims_list <- gen_sims_ppp(999) #simulations
sims_habitat_counts <- pbapply::pblapply(sims_list, FUN=compute_quadrat_counts) #apply counts function, with a progress bar from pbapply
full_count_df <- as_tibble(sims_habitat_counts) |> t() |> as_tibble() #smoosh them together
colnames(full_count_df) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies") #label columns
full_count_df |> head() #tada!

#this took two hours



```{r}
#write_csv(full_count_df, "data/raw/monte_carlo_sims.csv") #done after code above
#saveRDS(sims_list, "data/raw/sims_ppplist.rds")

#reading them back in after saving
monte_carlo_df <- read_csv('data/raw/monte_carlo_sims.csv')
barrows_sims_ppplist <- readRDS("data/raw/sims_ppplist.rds")

```


```{r}
#have to rename my categories actually so they work with the histograms
names(monte_carlo_df) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")

names(obs_barrow_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
```

```{r}
#| label: forests distribution
  mc_df <- bind_rows(monte_carlo_df, obs_barrow_counts)
  monte_carlo_df |> ggplot(aes(x=Forests)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Forests, color = "darkgreen") +
    theme_classic()
#hahahaha turns out that's way lower, so they actually AVOID forests. What a fun finding!

```

```{r}
#| label: dense human distribution
  monte_carlo_df |> ggplot(aes(x=Dense_Human_Habitat)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Dense_Human_Habitat, color = "red") +
    theme_classic()
#We're pretty close to the bounds for that one

```


```{r}
#| label: human agriculture
  monte_carlo_df |> ggplot(aes(x=Human_Agriculture)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Human_Agriculture, color = "orange") +
    theme_classic()
#WOW that's way higher, which is not what I was expecting. Will have to comment on this

```

```{r}
#| label: grass and shrublands
  monte_carlo_df |> ggplot(aes(x=Grass_and_Shrublands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Grass_and_Shrublands, color = "green") +
    theme_classic()
#We're pretty close to the bounds for that one too

```

```{r}
#| label: rocky, barren, and snowy
  monte_carlo_df |> ggplot(aes(x=Rocky_Barren_Snowy)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Rocky_Barren_Snowy) +
    theme_classic()
#Far below, they DONT like these spots

```

```{r}
#| label: wetlands
  monte_carlo_df |> ggplot(aes(x=Wetlands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Wetlands, color = "lightblue") +
    theme_classic()
#Far below, they DONT like these spots either

```

```{r}
#| label: water bodies 2
  monte_carlo_df |> ggplot(aes(x=Water_Bodies)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Water_Bodies, color = "blue") +
    theme_classic()
#Far below, they DONT like these spots either either

```


The p-value is generated by taking the proportion of the null distribution that has more fairy barrows predicted in that habitat type than the observed counts. In context of the Grass and Shrublands habitat, this means that 40% of the Monte Carlo simulations had more fairy barrows in Grass and Shrublands than the actual data. A p-value of 1.000 means that the true number of fairy barrows in that habitat type is far lower than the simulations, and indicates an avoidance behavior. A very small p-value, such as 0.001 in Human Agriculture, shows that a significantly higher proportion of fairy barrows were built in that habitat than would be expected, indicating a preference behavior. The distributions for each habitat type are displayed below, for comparison.

# P value calculations
```{r}
#| label: p value calculations

human_hab_count_more_extreme <- mc_df |> filter(Dense_Human_Habitat >= obs_barrow_counts$Dense_Human_Habitat)
human_hab_p_val <- nrow(human_hab_count_more_extreme) / nrow(mc_df)

ag_count_more_extreme <- mc_df |> filter(Human_Agriculture >= obs_barrow_counts$Human_Agriculture)
ag_p_val <- nrow(ag_count_more_extreme) / nrow(mc_df)

forest_count_more_extreme <- mc_df |> filter(Forests >= obs_barrow_counts$Forests)
forest_p_val <- nrow(forest_count_more_extreme) / nrow(mc_df)

grass_shrub_count_more_extreme <- mc_df |> filter(Grass_and_Shrublands >= obs_barrow_counts$Grass_and_Shrublands)
grass_shrub_p_val <- nrow(grass_shrub_count_more_extreme) / nrow(mc_df)

barren_count_more_extreme <- mc_df |> filter(Rocky_Barren_Snowy >= obs_barrow_counts$Rocky_Barren_Snowy)
barren_p_val <- nrow(barren_count_more_extreme) / nrow(mc_df)

wetland_count_more_extreme <- mc_df |> filter(Wetlands >= obs_barrow_counts$Wetlands)
wetland_p_val <- nrow(wetland_count_more_extreme) / nrow(mc_df)

water_count_more_extreme <- mc_df |> filter(Water_Bodies >= obs_barrow_counts$Water_Bodies)
water_p_val <- nrow(water_count_more_extreme) / nrow(mc_df)

#I could make this into a df instead of doing this way huh?

print("P-values by habitat:")
print(paste("Dense Human Habitat:",human_hab_p_val))
print(paste("Human Agriculture:",ag_p_val))
print(paste("Forest:",forest_p_val))
print(paste("Grass and Shrublands:",grass_shrub_p_val))
print(paste("Rocky, Barren, and Snowy:",barren_p_val))
print(paste("Wetlands:",wetland_p_val))
print(paste("Water Bodies:",water_p_val))

#so out of all of these only agriculture is truly significant, but all the '1'basically means the true count was WAY lower than the expected count, which is also significant, and I'll work on that in a bit
p_values <- c(human_hab_p_val, ag_p_val, forest_p_val, grass_shrub_p_val, barren_p_val, wetland_p_val,water_p_val)
p_val_greater_df <- data.frame(habitat = lc_labels, p_value = p_values)
print(p_val_greater_df)

```

From these distributions and their associated p-values, we find that fairies prefer Human Agriculture as a habitat, and avoid forests, barren areas, wetlands, and water bodies. They have no particular preference or aversion for Dense Human Habitat or Grass and Shrublands. Further studies must be conducted to study the reason for this new finding. Perhaps they prefer to hunt in forests and wetlands, which is why they are encountered so often but do not reside there. Perhaps our data was false, to conceal where they reside, or shows evidence of a misinformation campaign that we just uncovered. Regardless, no further conclusions can be drawn until a behavioral study is performed.

# Hypothesis 2: Proximity to Stone Circles
Stone circles are said to be portals between the fairy realm and our realm, and there have been theories that fairies use these portals regularly. If this were the case, the fairies may choose to live close to stone circles, to prevent frequent cross-country migration. In addition, ecological migration theories for other animal species indicate that fairies might emerge from a portal and choose a nearby location to construct their barrows.

Null Hypothesis: Fairies in the Republic of Ireland display no significant clustering in home locations, indicating no preference in habitat.

Hypothesis 2: Fairies in the Republic of Ireland prefer to live near stone circles.

To test this hypothesis, we will calculate the distance of each fairy barrow to its nearest stone circle. This will generate our observation statistic, and then we will run the same calcualtion on the 999 simulated barrow patterns from Hypothesis 1. This will create a null distribution of the average distance a barrow might be constructed from a stone circle. 

```{r}
#| label: calculating test statistic of nearest neighbor distances
obs_dists_sc <- nncross(barrows_ppp, stone_circles_ppp)$dist
mean_obs_dists <- mean(obs_dists_sc)
```

```{r}
mean_obs_dists
```

```{r}
#| label: creating the distance function for the whole dataset
compute_mean_dist <- function(sim_ppp) {
  sim_dists <- nncross(sim_ppp, stone_circles_ppp)$dist
  mean(sim_dists)
}

sim_distances <- lapply(X=barrows_sims_ppplist, FUN=compute_mean_dist)

```

```{r}
#| label: visualizing mean distance null distribution

sim_dist_df <- unlist(sim_distances) |> as_tibble()
obs_dist_df <- tibble(value = mean_obs_dists)
all_dists_df <- rbind(sim_dist_df,obs_dist_df)
all_dists_df |> ggplot(aes(x=value)) +
  geom_histogram(fill='black', alpha=0.5) +
  geom_vline(xintercept = mean_obs_dists, color='darkorchid4') +
  theme_classic()
```


```{r}
#| label: computing distance p value

count_more_extreme <- all_dists_df |> filter(value >= mean_obs_dists)
sc_dist_p_val <- nrow(count_more_extreme) / nrow(all_dists_df)
sc_dist_p_val
```

This is another important finding on fairy behavior and home preference. People have historically observed fairies emerging from stone circles, and many humans have gone missing upon visiting them. However, the relationship between fairy dwellings and their nearby portals has never been studied or quantified. Further studies should be completed on how often a fairy travels to a stone circle, and whether their frequency is at all associated with the distance of their home to the nearest portal. 

# Hypothesis 3: Stone Circle Prominent Habitats
Some modern researchers claim that stone circles are built in places that resemble the fairy realm on the other side. This is, as they say, part of the power that connects the realms and allows transportation between. Unfortunately, as seen in the literature review, very little evidence (and researchers) of these findings have survived. In our hubris, we seek to determine whether this might be true. If stone circles are prominently built in a certain habitat, it may give us insight into the nature of the fairy realm. If I go missing upon the publication of this paper, this hypothesis will be why.

Hypothesis 3: Stone circles were built in a certain habitat, possibly reflecting the nature of the fairy realm on the other side.

To test this hypothesis, we will run a very similar test to that of the fairy barrows. We will generate a habitat 'intensity' funcion and use it to determine whether stone circles are clustered in any of the major land cover types. The number of stone circles found in each habitat type will form the initial observations, which we will compare to a Monte Carlo simulation. The Monte Carlo simulation will generate 999 similar point patterns of hypothetical stone circles, proportionally generated into each habitat type. We will then compare the probability of a stone circle appearing in each habitat type to the true proportion of real circles we have observed. 

## Monte Carlo Simulations: Stone Circles

```{r}
#creating the stone circles function, similar to the barrows one
num_circles <- nrow(stone_circles_sf) #to set number of points to sample

#creating a function that will plot barrows around ireland according to the area proportions of the various habitats
#and checking the runtime to make sure we can do the full 999 sims
runtime <- system.time({gen_circles_sim <- function() {
  circles_sim <- rpoint(n=num_circles,f=habitat_prob_intensity, win=stone_circles_ppp$window)
  return(circles_sim) 
}
single_circles_sim_ppp <- gen_circles_sim()
#IT WORKED!!! YAYYYYYY
single_circle_sim_counts <- quadratcount(single_circles_sim_ppp, tess=lc_tess) |> as.vector()
names(single_circle_sim_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
single_circle_sim_counts <- single_circle_sim_counts |> as_tibble_row()})
single_circle_sim_counts
runtime

```

now to take a little while again


library('pbapply')
#Now we are running 999 simulations wahoooo

compute_quadrat_counts <- function(sim_ppp) { #making a function to apply to all the sims
  sim_counts <- quadratcount(sim_ppp, tess=lc_tess) |> as.vector()
  names(sim_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
  return(sim_counts)
}

gen_sims_ppp <- function(num_sims) { #function to generate 999 ppp simulations
  stone_circle_sims <- spatstat.random::rpoint(
    n = num_circles,
    f = habitat_prob_intensity,
    win = stone_circles_ppp$window,
    nsim = num_sims
  )
  return(stone_circle_sims)
}

sc_sims_list <- gen_sims_ppp(999) #simulations
sc_habitat_counts <- pbapply::pblapply(sc_sims_list, FUN=compute_quadrat_counts) #apply counts function, with a progress bar from pbapply
sc_count_df <- as_tibble(sc_habitat_counts) |> t() |> as_tibble() #smoosh them together
colnames(sc_count_df) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies") #label columns
sc_count_df |> head() #tada!

#this took two hours


```{r}
#saving so we dont have to do that again either
#write_csv(sc_count_df, "data/raw/sc_monte_carlo_sims.csv") #done after code above
sc_monte_carlo_df <- read_csv('data/raw/sc_monte_carlo_sims.csv')
```

Now to generate that test statistic, we are slightly out of order here

```{r}
obs_sc_counts <- quadratcount(stone_circles_ppp, tess=lc_tess) |> as.vector()
names(obs_sc_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
obs_sc_counts <- obs_sc_counts |> as_tibble_row()
```

```{r}
obs_sc_counts
```

And the distributions for all habitats

```{r}
#| label: forests distribution stone circles
  sc_df <- bind_rows(sc_monte_carlo_df, obs_sc_counts)
  sc_monte_carlo_df |> ggplot(aes(x=Forests)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Forests, color = "darkgreen") +
    theme_classic()
#right on target, so nothing special there

```

```{r}
#| label: dense human distribution stone circles
  sc_monte_carlo_df |> ggplot(aes(x=Dense_Human_Habitat)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Dense_Human_Habitat, color = "red") +
    theme_classic()
#also just about right

```


```{r}
#| label: human agriculture stone circles
  sc_monte_carlo_df |> ggplot(aes(x=Human_Agriculture)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Human_Agriculture, color = "orange") +
    theme_classic()
#right in the middle

```

```{r}
#| label: grass and shrublands stone circles
  sc_monte_carlo_df |> ggplot(aes(x=Grass_and_Shrublands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Grass_and_Shrublands, color = "green") +
    theme_classic()
#high for this one! nice!

```

```{r}
#| label: rocky, barren, and snowy stone circles
  sc_monte_carlo_df |> ggplot(aes(x=Rocky_Barren_Snowy)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Rocky_Barren_Snowy) +
    theme_classic()
#eh maybe higher, not a great distribution though

```

```{r}
#| label: wetlands stone circles
  sc_monte_carlo_df |> ggplot(aes(x=Wetlands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Wetlands, color = "lightblue") +
    theme_classic()
#in the middle

```

```{r}
#| label: water bodies
  sc_monte_carlo_df |> ggplot(aes(x=Water_Bodies)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Water_Bodies, color = "blue") +
    theme_classic()
#Far below, makes sense though, maybe put something about avoiding running water

```


# P value calculations for stone circles
```{r}
#| label: p value calculations for stone circle

human_hab_count_more_extreme_sc <- sc_df |> filter(Dense_Human_Habitat >= obs_sc_counts$Dense_Human_Habitat)
human_hab_p_val_sc <- nrow(human_hab_count_more_extreme_sc) / nrow(sc_df)

ag_count_more_extreme_sc <- sc_df |> filter(Human_Agriculture >= obs_sc_counts$Human_Agriculture)
ag_p_val_sc <- nrow(ag_count_more_extreme_sc) / nrow(sc_df)

forest_count_more_extreme_sc <- sc_df |> filter(Forests >= obs_sc_counts$Forests)
forest_p_val_sc <- nrow(forest_count_more_extreme_sc) / nrow(sc_df)

grass_shrub_count_more_extreme_sc <- sc_df |> filter(Grass_and_Shrublands >= obs_sc_counts$Grass_and_Shrublands)
grass_shrub_p_val_sc <- nrow(grass_shrub_count_more_extreme_sc) / nrow(sc_df)

barren_count_more_extreme_sc <- sc_df |> filter(Rocky_Barren_Snowy >= obs_sc_counts$Rocky_Barren_Snowy)
barren_p_val_sc <- nrow(barren_count_more_extreme_sc) / nrow(sc_df)

wetland_count_more_extreme_sc <- sc_df |> filter(Wetlands >= obs_sc_counts$Wetlands)
wetland_p_val_sc <- nrow(wetland_count_more_extreme_sc) / nrow(sc_df)

water_count_more_extreme_sc <- sc_df |> filter(Water_Bodies >= obs_sc_counts$Water_Bodies)
water_p_val_sc <- nrow(water_count_more_extreme_sc) / nrow(sc_df)


p_values_sc <- c(human_hab_p_val_sc, ag_p_val_sc, forest_p_val_sc, grass_shrub_p_val_sc, barren_p_val_sc, wetland_p_val_sc,water_p_val_sc)
p_val_greater_df_sc <- data.frame(habitat = lc_labels, p_value = p_values_sc)
print(p_val_greater_df_sc)

```

From the p-values, we find that roughly the expected number of stone circles exists in each habitat except for Grass and Shrublands, which has a disproportionately high number of stone circles. It also appears that stone circles are not built in water bodies, but that is not a surprise. Below, we see the null distribution of the likely number of stone cirlces in the Grass and Shrublands, shown in grey bars, and the true number of stone circles found, shown by a light green bar. If we had the 'expected' number of circles in Grass and Shrublands, the colored line would be somewhere within the grey null distribution. Instead it is to the far right side of the distribution, and 0.9% of the possible stone circle counds are higher than the true number of stone circle counts in that habitat. Collectively, this indicates that fairies prefer to build their stone circles in Grass and Shrubland habitats, but have no aversions beyond water bodies, which would be hard to build in. 

Since fairies prefer to build their stone circles in Grass and Shrublands, and tend to build their homes near these stone circles, we must ensure that Ireland is properly preserving its Grass and Shrublands. The following section will investigate this extension of the original hypotheses.

## Land Cover Change Analysis
Through our hypothesis testing, we determined that:
1) Fairies prefer to live in agricultural areas, defying centuries of literature

2) Fairies prefer to live in close proximity to stone circles

3) Stone circles are most commonly built in grasslands and shrublands

Because stone circles appear to be central to fairy migration and intentional in their construction, the Republic of Ireland should ensure they are preserving their grasslands and shrublands. All of the previous analysis was done on land cover data from 1990, which is the oldest land cover data available for the Republic of Ireland. The following section will compare the 1990 land cover data to the most recent land cover data, from 2018. We will specifically look at the Grass and Shrublands area, and create a transition matrix to see how much area was lost or gained over the intervening time. The Grass and Shrublands area makes up less than 1% of Irelands total land area, which means there is a very limited amount of land for stone circle construction. 

A change matrix is displayed below, with each cell representing the number of 100mx100m pixels that transformed into that habitat type between 1990 and 2018. The rows represent the original 1990 land cover type, while the columns indicate the 2018 land cover type. For example, cell (1,2) represents the 13014 pixels that changed from Dense Human Habitat to Human Agriculture in the intervening time.

Time to see if we have lost grass and shrubland area! Thankfully we already have the 1990 reclassified raster, and need to do the same thing for 2018

```{r}
#| label: land cover change analysis, grasslands
#reclassifying the 2018 data
rc_lc_2018 <- classify(lc_2018, rclmat, include.lowest=TRUE)
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")
plot(rc_lc_2018, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 2018")

#realigning the pixels to make sure they're on top of each other
rc_lc_2018 <- resample(rc_lc_2018, rc_lc_1990, method = "near")


#stacking the raster so the pixels are paired
lc_stack <- c(rc_lc_1990, rc_lc_2018)
names(lc_stack) <- c("1990", "2018")

#changing the column/row names
lc_class_names <- c(
  "1" = "Dense_Human_Habitat",
  "2" = "Human_Agriculture",
  "3" = "Forests",
  "4" = "Grass_and_Shrublands",
  "5" = "Rocky_Barren_Snowy",
  "6" = "Wetlands",
  "7" = "Water_Bodies"
)

change_matrix <- crosstab(lc_stack, long = FALSE)
colnames(change_matrix) <- lc_class_names[colnames(change_matrix)]
rownames(change_matrix) <- lc_class_names[rownames(change_matrix)]

change_matrix

```

This change matrix represents the land cover transitions for every habitat type. To determine the gain or loss in Grass and Shrublands, we will complete a second analysis. The table below shows the number of pixels that were Grass and Shrublands in both time periods (Persistent), the number of pixels that turned from another habitat type into Grass and Shrublands (Gain), the number of pixels that turned from Grass and Shrublands to a different habitat type (Loss), and the number of other pixels in Ireland (Never Grass). 

```{r}
# we know our grass and shrublands are class 4
grass_code <- 4

#this got complicated quick, but basically I had to make an ireland-shaped polygon of nothing, with pixels where grassland exists
grass_1990_r <- rc_lc_1990
values(grass_1990_r) <- ifelse(is.na(values(rc_lc_1990)), NA,
                               values(rc_lc_1990) == grass_code)

grass_2018_r <- rc_lc_2018
values(grass_2018_r) <- ifelse(is.na(values(rc_lc_2018)), NA,
                               values(rc_lc_2018) == grass_code)

# now I have to make a little transition raster that's also an empty ireland-shaped polygon of nothing
transition <- rc_lc_1990
values(transition) <- NA

# and make sure we only update the cells that are inside the original land cover raster (this got so technical)
valid <- !is.na(values(rc_lc_1990))

#now I'm making my new transition classes all about grassland
values(transition)[valid & values(grass_1990_r) & values(grass_2018_r)] <- 1  #stayed as grassland
values(transition)[valid & !values(grass_1990_r) & values(grass_2018_r)] <- 2 #new grassland
values(transition)[valid & values(grass_1990_r) & !values(grass_2018_r)] <- 3 #lost grassland
values(transition)[valid & !values(grass_1990_r) & !values(grass_2018_r)] <- 0 #never grassland

#and plotting
transition_colors <- c("grey80","darkgreen","lightgreen","saddlebrown")
plot(transition, col = transition_colors, type="classes",
     main="Grass and Shrublands Transitions", axes=TRUE)

```

And I want the real numbers of the transition, so we're going back to the change matrix to grab those

```{r}
persistent <- change_matrix["Grass_and_Shrublands","Grass_and_Shrublands"]
gain <- sum(change_matrix[-which(rownames(change_matrix)=="Grass_and_Shrublands"),
                         "Grass_and_Shrublands"]) #anything that didnt have grass in 1990 but does in 2018

loss <- sum(change_matrix["Grass_and_Shrublands",
                          -which(colnames(change_matrix)=="Grass_and_Shrublands")]) #grass in 1990 but not in 2018

never_grass <- sum(change_matrix[-which(rownames(change_matrix)=="Grass_and_Shrublands"),
                               -which(colnames(change_matrix)=="Grass_and_Shrublands")]) #never grass

grass_summary <- c(Persistent = persistent,
                   Gain = gain,
                   Loss = loss,
                   Never_Grass = never_grass)
print(grass_summary)

#and checking that the raster has the proper pixels, which it does!
pixel_counts <- freq(transition)

print(pixel_counts)


```

From the above table, we see that the Republic of Ireland saw a 12.3% increase in total Grass and Shrubland area from 1990 to 2018. About half of the original grassland from 1990 was lost, but a far greater amount of new grassland formed by 2018. While this does provide opportunities for the fairies to create new stone circles, it does imply that some of the original stone circles might have been destroyed or shifted into different habitat types. Further research must be conducted to see if the Loss grasslands included a large amount of stone circles, or whether the areas with stone circles were in the half that was preserved.

# Conclusion
Much to the author's relief, this concluses the initial investigation of fairy habitats in the Republic of Ireland. This study provides evidence that, despite popular belief, Irish fairies prefer to live in agricultural areas, rather than the oft-depicted forests. In addition, fairies prefer to live near stone circles, possibly to easily travel to them, or as a result of their initial dispersal to the human realm via the portals. Finally, we found that stone circles are disproportionately often constructed in grasslands and shrublands. Upon investigation of land cover change, we found that the Republic of Ireland has maintained and increased the total amount of available grasslands for possible stone circle construction, but that half of the grasslands since 1990 have transitioned to other land cover types. We recommend immediate research into the effects of this loss, and whether any major fairy circles were disturbed between 1990 and 2018. We also recommend behavioral studies on the preference for agricultural dwellings, and propose research into the hunting habits of Irish fairies as an explanation for their frequent sightings in forested areas. Finally, we offer advice to future researchers: Offerings of honey, flowers, and interesting buttons may sate the fairies long enough to complete your studies.