---
title: Kylie King's DSAN 6750 GIS project
authors:
  - name: Kylie King
    affiliation: Georgetown University
    roles: writing
    corresponding: true
bibliography: references.bib
---

## Abstract
The purpose of this project is to investigate whether Ireland is caring for its historic fairy habitats. Because fairies are notorious for obstructing data collection and interfering with computer calculations, there is very little data available on fairy habitats and their preservation. This is a two-part project, where we first determine where fairies prefer to live, and then assess whether their preferred habitat is flourishing. Data collected (with much sacrifice) on the location of fairy barrows (homes) will be converted into a density layer to determine where fairy barrows are most densely clustered. Next, land cover data from 1990 will be compared to the barrow density to determine which habitat fairies prefer. Finally,  land cover data from 2018 will be compared to the 1990 dataset, to produce a change matrix, and identify whether Irish fairy habitat remains intact.


## Literature Review
mostly talking about stories about fairies, for funsies

# Data Collection and Preparation
Barrows data, two land cover datas, natural earth data country layer

```{r}
#| label: initializing
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
set.seed(6805)
```

```{r}
#| label: data preparation

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, county, smrs, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

```{r}
#| label: getting ireland shapefile

ireland <- ne_countries(country = "ireland", scale="large")
mapview(ireland)
```

ok so, instead of the intensity function of population which is divided into low, med, and high density, I have the raster dataset which is split into ~5 habitat types

## Hypotheses
Hypothesis 1: Fairies mainly live in forested areas
Hypothesis 2: In accordance with global trends, fairy habitat is declining.

## Where Do Fairies Live?
probably forests I think


# Viewing Window Construction
```{r}
#| label: viewing window construction
ireland_sfc <- ireland |> st_convex_hull() #unsure if we will need this yet, or if the base shapefile will work for all the points

ireland <- ireland |> sf::st_transform(2157)

ireland_sfc <- ireland_sfc |> sf::st_transform(2157) #Irish Transverse Mercator
mapview(ireland_sfc)

```

# Intensity Layer Construction
wahoooo actually do i need this? Gonna read in the raster first to see what's up

# Land Cover Analysis
this is where I have to do interesting things with a raster, not sure how to do that yet. Will definitely have to reclassify to make sure the two land covers match

```{r}
#| label: reading in the rasters

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)

# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)

```

```{r}
#| label: visualizing 1990 raster
# This works!!
# Convert the raster to a data frame for ggplot2
lc_df_test <- as.data.frame(reprojected_1990, xy=TRUE) |>
  rename(value = U2000_CLC1990_V2020_20u1) # Rename the value column

# Ensure the 'value' column is treated as a factor for discrete colors
lc_df_test$value <- factor(lc_df_test$value)

# Plot using ggplot2
ggplot(data = lc_df_test, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  labs(title = "Raster Symbology by Unique Values", fill = "Category") +
  theme_minimal()
```

```{r}
#| label: reclassifying
#plot(lc_2018)
```


## Fairy Habitat Preservation
small little paragraph about what we have to do, also maybe prepping the land cover

# Change Matrix
no idea how do to this, wheeeeeeee

## Discussion and Findings
the irish people are probably in danger because i bet fairy habitat has declined