---
title: Kylie King's DSAN 6750 GIS project
authors:
  - name: Kylie King
    affiliation: Georgetown University
    roles: writing
    corresponding: true
bibliography: references.bib
---

## Abstract

The purpose of this project is to investigate whether Ireland is caring for its historic fairy habitats. Because fairies are notorious for obstructing data collection and interfering with computer calculations, there is very little data available on fairy habitats and their preservation. This is a two-part project, where we first determine where fairies prefer to live, and then assess whether their preferred habitat is flourishing. Data collected (with much sacrifice) on the location of fairy barrows (homes) will be converted into a density layer to determine where fairy barrows are most densely clustered. Next, land cover data from 1990 will be compared to the barrow density to determine which habitat fairies prefer. Finally, land cover data from 2018 will be compared to the 1990 dataset, to produce a change matrix, and identify whether Irish fairy habitat remains intact.

## Literature Review

mostly talking about stories about fairies, for funsies

# Data Collection and Preparation

Barrows data, two land cover datas, natural earth data country layer

```{r}
#| label: initializing
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
set.seed(6805)
```

```{r}
#| label: data preparation

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, county, smrs, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

```{r}
#| label: getting ireland shapefile

ireland <- ne_countries(country = "ireland", scale="large")
mapview(ireland)
```

ok so, instead of the intensity function of population which is divided into low, med, and high density, I have the raster dataset which is split into \~5 habitat types

## Hypotheses

Hypothesis 1: Fairies mainly live in forested areas Hypothesis 2: In accordance with global trends, fairy habitat is declining.

## Where Do Fairies Live?

probably forests I think

# Viewing Window Construction

```{r}
#| label: viewing window construction
ireland_sfc <- ireland |> st_convex_hull() #unsure if we will need this yet, or if the base shapefile will work for all the points

ireland <- ireland |> sf::st_transform(2157)

ireland_sfc <- ireland_sfc |> sf::st_transform(2157) #Irish Transverse Mercator
mapview(ireland_sfc)

```

# Intensity Layer Construction

wahoooo actually do i need this? Gonna read in the raster first to see what's up

# Land Cover Analysis

this is where I have to do interesting things with a raster, not sure how to do that yet. Will definitely have to reclassify to make sure the two land covers match

```{r}
#| label: reading in the rasters

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)


# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)

```

```{r}
#| label: visualizing 1990 raster
# This works!!
# Convert the raster to a data frame for ggplot2
lc_df_test <- as.data.frame(lc_1990, xy=TRUE) |>
  rename(value = U2000_CLC1990_V2020_20u1) # Rename the value column

# Ensure the 'value' column is treated as a factor for discrete colors
lc_df_test$value <- factor(lc_df_test$value)

# Plot using ggplot2
ggplot(data = lc_df_test, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  labs(title = "Raster Symbology by Unique Values", fill = "Category") +
  theme_minimal()
```

```{r}
#| label: reclassifying
rclmat <- matrix(c(1, 11, 1,
                   11.999, 22, 2,
                   22.999, 25, 3,
                   25.999, 27, 4,
                   27.999, 29, 3,
                   29.999, 34, 5,
                   34.999, 39, 6,
                   39.999, 44, 7
                   ), ncol=3, byrow=TRUE)

rc_test_1990 <- classify(lc_1990, rclmat, include.lowest=TRUE)
```

```{r}
#| label: reclassifying visualization
rc_test_1990_df <- as.data.frame(rc_test_1990, xy=TRUE) |>
  rename(value = U2000_CLC1990_V2020_20u1) # Rename the value column

# Ensure the 'value' column is treated as a factor for discrete colors
rc_test_1990_df$value <- factor(rc_test_1990_df$value)

# Plot using ggplot2
ggplot(data = rc_test_1990_df, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  labs(title = "Reclassified Land Cover Values") +
  scale_fill_manual(
    name = "Reclassified Land Cover", 
    labels = c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies"),
    values = c("red3", "wheat2", "darkgreen", "darkolivegreen1","bisque4","aquamarine1","dodgerblue2")) +
  theme_minimal()
```

```{r}
#| label: turning rasters into sf object to see what happens
# 1. Create a sample SpatRaster

crs(rc_test_1990) <- "epsg:4326" # Assign an initial CRS (e.g., WGS84)

# 2. Convert the SpatRaster to SpatVector (polygons)
test_polygons_terra <- as.polygons(rc_test_1990)

# 3. Convert the SpatVector to an sf object
test_polygons_sf <- st_as_sf(test_polygons_terra)

# You can now work with 'polygons_sf' as a standard sf object.
mapview(test_polygons_sf)
#that worked perfectly! Now we have our 'zones' and can do a monte carlo simulation!
```

## Fairy Habitat Preservation

small little paragraph about what we have to do, also maybe prepping the land cover

# Change Matrix

no idea how do to this, wheeeeeeee

## Discussion and Findings

the irish people are probably in danger because i bet fairy habitat has declined

## Starting from the beginning, to get some structure

Our 3 datasets: - Point data of fairy barrows, to determine if there is clustering in particular habitats - Point data of stone circles, to see if THOSE are in any particular habitat (because they're bridges between worlds, so need to be in similar places) - Shapefile from natural earth data of ireland, at highest resolution - Raster data of land cover from 1990, to identify the habitats (and create the polygons to test for clustering) - Raster data of land cover from 2018, to determine if fairy habitat has shrunk significantly

Load all the libraries

```{r}
#| label: init 2
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
set.seed(6805)
```

Get the barrows point data into an sf object and visualize

```{r}
#| label: data preparation 2

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

Get the stone circles point data into an sf object, combine with barrows

```{r}
#| label: data preparation 3

stone_circles_df <- read_csv('data/raw/stone-circles.csv')
stone_circles_df <- stone_circles_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "stone circle")

stone_circles_sf <- stone_circles_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(stone_circles_sf,100)) #showing the first 100 points
head(stone_circles_sf)

```

```{r}
#| label: data preparation 4
fairies_sf <- rbind(barrows_sf, stone_circles_sf)
fairies_sf <- fairies_sf |> sf::st_transform(2157)

combined_map <- mapview(fairies_sf, zcol="type")

```

Get the polygon shapefile of ireland for the viewing window

```{r}
#| label: getting ireland shapefile 2

ireland <- ne_countries(country = "ireland", scale="large") |> select(name,geometry)
mapview(ireland)

ireland <- ireland |> sf::st_transform(2157)
mapview(ireland) + combined_map

```

Transform the fairy monuments file into a ppp object and visualize inside of ireland shapefile

```{r}
#| label: barrows ppp
barrows_sf <- barrows_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
barrows_sf <- barrows_sf[st_within(barrows_sf, ireland, sparse = FALSE), ] 
barrows_ppp <- as.ppp(barrows_sf, W=as.owin(ireland)) #ah, we do have to do a little bit of convex hull
barrows_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

```{r}
#| label: stone circles ppp
stone_circles_sf <- stone_circles_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
stone_circles_sf <- stone_circles_sf[st_within(stone_circles_sf, ireland, sparse = FALSE), ] 
stone_circles_ppp <- as.ppp(stone_circles_sf, W=as.owin(ireland)) #ah, we do have to do a little bit of convex hull
stone_circles_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

INSTEAD of using a population intensity function for the simulations, will use an area calculation. We will see if a higher proportion of fairies lives in a habitat than would be expected for the area of that habitat

which means now we bring in the rasters!

```{r}
#| label: reading in the rasters 2

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)
#lc_1990_stars <- read_stars(tiff_file_path_1990) #didn't work
#lc_1990 <- terra::project(lc_1990, "epsg:2157") #projecting to ITM


# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)
#lc_2018_stars <- read_stars(tiff_file_path_2018) #didn't work
#lc_2018 <- terra::project(lc_2018, "epsg:2157") #projecting to ITM


```

```{r}
#| label: reclassification 1990 2
rclmat <- matrix(c(1, 11, 1,
                   11.999, 22, 2,
                   22.999, 25, 3,
                   25.999, 27, 4,
                   27.999, 29, 3,
                   29.999, 34, 5,
                   34.999, 39, 6,
                   39.999, 44, 7
                   ), ncol=3, byrow=TRUE) #decimals so that it catches all possible values

rc_lc_1990 <- classify(lc_1990, rclmat, include.lowest=TRUE)

```

```{r}
#| label: visualizing the lc raster 1990
#creating the color library for the classes
lc_colors <- data.frame(
  value = c(1, 2, 3, 4, 5, 6, 7),
  color = c("#8B0000", "#F5DEB3", "#006400", "#556B2F","#8B7355", "#7FFFD4", "#1E90FF")
)

lc_labels <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")

plot(lc_1990, type = "classes", main = "Land Cover: 1990")
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")

```

Full land cover reclassification table available in data README. Original land cover had 44 discrete classes, which was reduced to 7 discrete classes.

| New Land Cover Category | New Cell Value |
|----------|------|
| Dense Human Habitat | 1 |
| Human Agriculture | 2 |
| Forests | 3 |
| Grass and Shrublands | 4 |
| Rocky, Barren, or Snowy | 5 |
| Wetlands | 6 |
| Water Bodies | 7 |


Our second hypothesis will be average distance from fairy circles AND what habitats those fairy circles were built in (using a similar simmulation probably)

Turning the classified raster into polygons so we can start doing the simulations

```{r}
#| label: raster into polygons
lc_poly_1990 <- as.polygons(rc_lc_1990)

# turning it into an sf object
lc_poly_sf <- st_as_sf(lc_poly_1990)

#dissolving the polygons so we can do an area calculation
lc_poly_sf <- lc_poly_sf |>
  group_by(U2000_CLC1990_V2020_20u1) |>
  summarise()

lc_poly_sf <- st_transform(lc_poly_sf, 2157)
lc_poly_sf <- st_intersection(lc_poly_sf, ireland) #clipping the polygons so the borders are the same as my viewing window

mapview(lc_poly_sf, zcol = "U2000_CLC1990_V2020_20u1")

```



Generating test statistic!
```{r}
#| label: barrow counts by habitat
#making lc into a tesselation
lc_list <- split(lc_poly_sf, lc_poly_sf$U2000_CLC1990_V2020_20u1) #grabbing individual polygons
lc_tiles <- lapply(lc_list, as.owin) #each one as a window

#tesselation of the polygon windows
lc_tess <- tess(tiles = lc_tiles)

#now the test statistic
obs_barrow_counts <- quadratcount(barrows_ppp, tess=lc_tess) |> as.vector()
names(obs_barrow_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
obs_barrow_counts <- obs_barrow_counts |> as_tibble_row()
```

The test statistic: Referencing the literature, it is believed that fairies prefer forested habitats, which offer readily available hollow trees, babbling streams, and mossy grottos. Our first hypothesis, therefore, is that fairies prefer to make their homes in forests.

```{r}
obs_barrow_counts
```


habitat 'intensity', which gives the proportional probability of any point existing in each habitat type (based on area)
```{r}
#| label: habitat intensity function
tile_areas <- sapply(lc_tiles, area.owin) #the area of each tile
tile_probs <- tile_areas / sum(tile_areas) #getting the proportions

habitat_prob_intensity <- function(x, y) { #intensity function for probability of points being in each habitat
  sapply(seq_along(lc_tiles), function(i) {
    inside.owin(x, y, lc_tiles[[i]]) * tile_probs[i]
  }) |> rowSums()
}
```


## Monte Carlo Simulations
```{r}
num_barrows <- nrow(barrows_sf)

#creating a function that will plot barrows around ireland according to the area proportions of the various habitats
gen_barrows_sim <- function() {
  barrows_sim <- rpoint(n=num_barrows,f=habitat_prob_intensity, win=barrows_ppp$window)
  return(barrows_sim) 
}

single_barrows_sim_ppp <- gen_barrows_sim()
single_barrows_sim_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

#IT WORKED!!! YAYYYYYY
single_barrow_sim_counts <- quadratcount(single_barrows_sim_ppp, tess=lc_tess) |> as.vector()
names(single_barrow_sim_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
single_barrow_sim_counts <- single_barrow_sim_counts |> as_tibble_row()
single_barrow_sim_counts
```
