---
title: Kylie King's DSAN 6750 GIS project
authors:
  - name: Kylie King
    affiliation: Georgetown University
    roles: writing
    corresponding: true
bibliography: references.bib
---

## Abstract

The purpose of this project is to investigate whether Ireland is caring for its historic fairy habitats. Because fairies are notorious for obstructing data collection and interfering with computer calculations, there is very little data available on fairy habitats and their preservation. This is a two-part project, where we first determine where fairies prefer to live, and then assess whether their preferred habitat is flourishing. Data collected (with much sacrifice) on the location of fairy barrows (homes) will be converted into a density layer to determine where fairy barrows are most densely clustered. Next, land cover data from 1990 will be compared to the barrow density to determine which habitat fairies prefer. To test whether fairies cluster near stone circles, rather than a particular habitat, a  will be done on the average distance of fairy barrows from nearby stone circles. The preferred habitat for stone circles will also be noted. Finally, land cover data from 2018 will be compared to the 1990 dataset, to produce a change matrix, and identify whether Irish fairy habitat remains intact.

## Introduction
Ireland's human inhabitants share the land with a cryptic, fickle, whimsical species known as the Aes Sídhe. They are also known as the Wee Folk, the Good People, the Fair Folk, or,  most generally, fairies. The ancient name, Aes Sídhe, means *folk of the fairy mounds,* referring to the mounds the fairies live in. These mounds, also known as *barrows,* are found throughout Ireland, and are regarded with superstition and reverence. This caution is well-founded, as those who disturb or damage these mounds are punished with extensive bad luck and guaranteed disaster. 

Unfortunately, the secreteive (and vengeful) nature of the Folk makes them difficult to study. As a result, modern researchers know very little about their preferred habitats, and the growing climate crisis is leading to worries about habitat suitability for Ireland's Good Neighbors (another moniker). Restless fairies are prone to violent outbursts, which presents a danger to Ireland's human inhabitants. 

This study will use the little data we have to determine if fairies have a preferred habitat, and extend that finding to assess whether Ireland is properly preserving that preferred habitat. If the author survives to the completion of this study, it may offer important insights about the future of humanity's relationship with our fairy neighbors. 

The hypotheses tested in this study are as follows: (ok so we need a way to say "yeah this was proven wrong, so we pivoted!")

Null Hypothesis: Fairies in the Republic of Ireland display no significant clustering in home locations, indicating no preference in habitat.

Hypothesis 1: Fairies in the Republic of Ireland prefer forested habitats.

Hypothesis 2: Fairies in the Republic of Ireland prefer to live near stone circles.

Hypothesis 3: Stone circles were built in a certain habitat, reflecting the nature of the fairy realm on the other side.

(fairies prefer to live in separate habitats than their hunting habitats, indicating a deep misunderstanding in fairy ecology)


# Literature Review

mostly talking about stories about fairies, for funsies
things to establish:
- We're only talking about the republic of ireland because, while fairies do not recognize international borders, the republic of ireland and northern ireland have different relationships with their respective Fair Folk
- we think Irish fairies will prefer forest habitats
- barrows data = fairy homes
- we also think they might prefer homes that are closer to stone circles
- stone circles data = portals to the ethereal realm or whatever its called in irish folklore
- fairies will be mad if their habitats degrade
- there is evidence that they may tamper with our data, which should be anticiapted
- There is also evidence that stories about fairies may be misleading
- for the nonbelievers, which will actually go at the end

barrows: "Barrows are typically defined as circular enclosures of earth and/or stone, often featuring a bank and ditch, though there are also examples without these features, consisting only of a simple mound. The interior of a barrow can take the form of a flat platform or shallow depression, but the most common type found in Ireland is the ring barrow. These barrows feature a low mound that rises gradually from the enclosing ditch and have an overall diameter of 15 to 25 meters."

stone circles: "These circles are characterised by the presence of standing stones, typically made of sandstone or quartz, that are arranged in a circular pattern. Stone circles can range in size from a few metres to over 100 metres in diameter, and they are thought to have served a variety of religious, cultural, and astronomical purposes."

## Methodology
To test our hypotheses, the locations of fairy barrows will be compared to land cover data from 1990 to determine whether fairies prefer a certain habitat (first-order clustering). To test whether fairies cluster near stone circles (second-order clustering), rather than a particular habitat, we will compare the average distance of fairy barrows to nearby stone circles. The preferred habitat for stone circle construction will also be tested with the same land cover data, as we have very little information about fairy habits. If any preferred habitats are found, land cover data from 2018 will be compared to the 1990 dataset to produce a change matrix, and identify whether fairy habitat in the Repulic of Ireland remains intact. 

# Study Environment Preparation

This study was conducted using R, with the packages tidyverse, sf, mapview, spatstat, stars, rnaturalearth, terra, tidyterra, and pbapply.

The Coordinate Reference System (CRS) for this study is the Irish Transverse Mercator (ITM) projection, with EPSG code 2157. All data is reprojected or transformed into this CRS for calculations and visualizations. (cite)

```{r}
#| label: itinialiigjgjg
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
library(pbapply) |> suppressPackageStartupMessages()
set.seed(6805)
```

# Fairy Point Data Preparation
Barrows and Stone Circles data were collected from the Irish Archaeology website, downloaded in kml format from the associated Interactive Google Map, and converted to CSV. The CSV was also cleaned in Microsoft Excel to retain all downloaded information in a usable format for the sf library. (cite) Metadata for these datasets can be found in the Appendix(link?). 

```{r}
#| label: barrows data prep

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

```{r}
#| label: circles data prep

stone_circles_df <- read_csv('data/raw/stone-circles.csv')
stone_circles_df <- stone_circles_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "stone circle")

stone_circles_sf <- stone_circles_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(stone_circles_sf,100)) #showing the first 100 points
head(stone_circles_sf)

```

```{r}
#| label: combiningn them for some reason, probably visualization help
fairies_sf <- rbind(barrows_sf, stone_circles_sf)
fairies_sf <- fairies_sf |> sf::st_transform(2157)

fairies_map <- mapview(fairies_sf, zcol="type")

```

The observation window for this study is the Natural Earth Data Republic of Ireland sf object at 1:10m resolution. All calculations are constrained to the bounds of this polygon. (cite)

```{r}
#| label: ireland observation window

ireland <- ne_countries(country = "ireland", scale="large") |> select(name,geometry)
ireland <- ireland |> sf::st_transform(2157)
mapview(ireland) + fairies_map

```

Each of the point layers were converted into a spatstat ppp (planar point pattern) object and plotted inside the Ireland observation window for the remainder of the calculations.

```{r}
#| label: barrows ppp 1
barrows_sf <- barrows_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
barrows_sf <- barrows_sf[st_within(barrows_sf, ireland, sparse = FALSE), ] 
barrows_ppp <- as.ppp(barrows_sf, W=as.owin(ireland)) 
barrows_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

```{r}
#| label: stone circles ppp 1
stone_circles_sf <- stone_circles_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
stone_circles_sf <- stone_circles_sf[st_within(stone_circles_sf, ireland, sparse = FALSE), ] 
stone_circles_ppp <- as.ppp(stone_circles_sf, W=as.owin(ireland)) 
stone_circles_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic()

```

# Land Cover Data Preparation

Land cover data from 1990 and 2018 was collected from the EU Copernicus component of the European Union's space program. It is produced by the Coordination of Information on the Environment (CORINE) program, which provides a uniform, continent-wide land cover inventory with 44 thematic classes called the CORINE Land Cover product. (cite a few things) Full metadata of the 44 classes and their associated reclassification can be found in the Appendix(link?).

```{r}
#| label: reading in the rasters againnnn

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)



# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)



```

```{r}
#| label: reclassification 1990 2 at the beginning
rclmat <- matrix(c(1, 11, 1,
                   11.999, 22, 2,
                   22.999, 25, 3,
                   25.999, 27, 4,
                   27.999, 29, 3,
                   29.999, 34, 5,
                   34.999, 39, 6,
                   39.999, 44, 7
                   ), ncol=3, byrow=TRUE) #decimals so that it catches all possible values

rc_lc_1990 <- classify(lc_1990, rclmat, include.lowest=TRUE)

```

```{r}
#| label: visualizing the lc raster 1990 at teh beginning (make pretty later)
#creating the color library for the classes
lc_colors <- data.frame(
  value = c(1, 2, 3, 4, 5, 6, 7),
  color = c("#8B0000", "#F5DEB3", "#006400", "#556B2F","#8B7355", "#7FFFD4", "#1E90FF")
)

lc_labels <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")

plot(lc_1990, type = "classes", main = "Land Cover: 1990")
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")

```

Full land cover reclassification table available in the Appendix(link?). Original land cover had 44 discrete classes, which was reduced to the following 7 discrete classes.

| New Land Cover Category | New Cell Value |
|----------|------|
| Dense Human Habitat | 1 |
| Human Agriculture | 2 |
| Forests | 3 |
| Grass and Shrublands | 4 |
| Rocky, Barren, or Snowy | 5 |
| Wetlands | 6 |
| Water Bodies | 7 |

## Hypothesis Testing
The following section has the details for each hypothesis test (we probably need to talk more about methods? Possibly)

# Hypothesis 1: Barrow Preferred Habitats

# Hypothesis 2: Proximity to Stone Circles

# Hypothesis 3: Stone Circle Prominent Habitats

## Land Cover Change Analysis

## Results and Discussion

## Appendix


## Walkthrough with lots of context and no polish

Our 3 datasets: - Point data of fairy barrows, to determine if there is clustering in particular habitats - Point data of stone circles, to see if THOSE are in any particular habitat (because they're bridges between worlds, so need to be in similar places) - Shapefile from natural earth data of ireland, at highest resolution - Raster data of land cover from 1990, to identify the habitats (and create the polygons to test for clustering) - Raster data of land cover from 2018, to determine if fairy habitat has shrunk significantly

Load all the libraries

```{r}
#| label: init 2
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(spatstat) |> suppressPackageStartupMessages()
library(stars) |> suppressPackageStartupMessages()
library(rnaturalearth) |> suppressPackageStartupMessages()
library(terra) |> suppressPackageStartupMessages()
library(tidyterra) |> suppressPackageStartupMessages()
library(pbapply) |> suppressPackageStartupMessages()
set.seed(6805)
```


Get the barrows point data into an sf object and visualize

```{r}
#| label: data preparation 2

barrows_df <- read_csv('data/raw/barrows.csv')
barrows_df <- barrows_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "barrow")

barrows_sf <- barrows_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(barrows_sf,100)) #showing the first 100 points
head(barrows_sf)

```

Get the stone circles point data into an sf object, combine with barrows

```{r}
#| label: data preparation 3

stone_circles_df <- read_csv('data/raw/stone-circles.csv')
stone_circles_df <- stone_circles_df |> select(townland, class_code, latitude, longitude) |> mutate(type = "stone circle")

stone_circles_sf <- stone_circles_df |> sf::st_as_sf(
  coords=c("longitude","latitude"),
  crs=4326) #projected to work with mapview, will reproject to Irish Transverse Mercator (ITM) (EPSG:2157)
  #also the land cover data is epsg 4326

mapview(head(stone_circles_sf,100)) #showing the first 100 points
head(stone_circles_sf)

```

```{r}
#| label: data preparation 4
fairies_sf <- rbind(barrows_sf, stone_circles_sf)
fairies_sf <- fairies_sf |> sf::st_transform(2157)

combined_map <- mapview(fairies_sf, zcol="type")

```

Get the polygon shapefile of ireland for the viewing window

```{r}
#| label: getting ireland shapefile 2

ireland <- ne_countries(country = "ireland", scale="large") |> select(name,geometry)
mapview(ireland)

ireland <- ireland |> sf::st_transform(2157)
mapview(ireland) + combined_map

```

Transform the fairy monuments file into a ppp object and visualize inside of ireland shapefile

```{r}
#| label: barrows ppp
barrows_sf <- barrows_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
barrows_sf <- barrows_sf[st_within(barrows_sf, ireland, sparse = FALSE), ] 
barrows_ppp <- as.ppp(barrows_sf, W=as.owin(ireland)) 
barrows_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf(size = 1) +
    theme_classic() +
    labs(title = "Fairy Barrows") + theme(
  plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )


```

```{r}
#| label: stone circles ppp
stone_circles_sf <- stone_circles_sf |> sf::st_transform(2157)
#some of the points are outside my ireland shapefile, so I am going to remove them (since I have thousands)
stone_circles_sf <- stone_circles_sf[st_within(stone_circles_sf, ireland, sparse = FALSE), ] 
stone_circles_ppp <- as.ppp(stone_circles_sf, W=as.owin(ireland)) 
stone_circles_ppp |> sf::st_as_sf() |> ggplot() +
    geom_sf() +
    theme_classic() +     
    labs(title = "Stone Circles") + theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )

```

INSTEAD of using a population intensity function for the simulations, will use an area calculation. We will see if a higher proportion of fairies lives in a habitat than would be expected for the area of that habitat

which means now we bring in the rasters!

```{r}
#| label: reading in the rasters 2

temp_dir <- tempdir() #temporary directory for working
zip_file_path <- "data/raw/Rasters.zip" #path to raster
unzip(zipfile = zip_file_path, exdir = temp_dir) #unzipping into temp directory

# 1990 dataset
tiff_file_path_1990 <- file.path(temp_dir, "Rasters/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1/U2000_CLC1990_V2020_20u1.tif")
lc_1990 <- rast(tiff_file_path_1990)
#lc_1990_stars <- read_stars(tiff_file_path_1990) #didn't work
#lc_1990 <- terra::project(lc_1990, "epsg:2157") #projecting to ITM


# 2018 dataset
tiff_file_path_2018 <- file.path(temp_dir, "Rasters/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1/U2018_CLC2018_V2020_20u1.tif")
lc_2018 <- rast(tiff_file_path_2018)
#lc_2018_stars <- read_stars(tiff_file_path_2018) #didn't work
#lc_2018 <- terra::project(lc_2018, "epsg:2157") #projecting to ITM


```

```{r}
#| label: reclassification 1990 2
rclmat <- matrix(c(1, 11, 1,
                   11.999, 22, 2,
                   22.999, 25, 3,
                   25.999, 27, 4,
                   27.999, 29, 3,
                   29.999, 34, 5,
                   34.999, 39, 6,
                   39.999, 44, 7
                   ), ncol=3, byrow=TRUE) #decimals so that it catches all possible values

rc_lc_1990 <- classify(lc_1990, rclmat, include.lowest=TRUE)

```

```{r}
#| label: visualizing the lc raster 1990
#creating the color library for the classes
lc_colors <- data.frame(
  value = c(1, 2, 3, 4, 5, 6, 7),
  color = c("#8B0000", "#F5DEB3", "#006400", "#99ec0aff","#8B7355", "#7FFFD4", "#1E90FF")
)

lc_labels <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")

plot(lc_1990, type = "classes", main = "Land Cover: 1990")
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")

```

Full land cover reclassification table available in data README. Original land cover had 44 discrete classes, which was reduced to 7 discrete classes.

| New Land Cover Category | New Cell Value |
|----------|------|
| Dense Human Habitat | 1 |
| Human Agriculture | 2 |
| Forests | 3 |
| Grass and Shrublands | 4 |
| Rocky, Barren, or Snowy | 5 |
| Wetlands | 6 |
| Water Bodies | 7 |


Our second hypothesis will be average distance from fairy circles AND what habitats those fairy circles were built in (using a similar simmulation probably)

Turning the classified raster into polygons so we can start doing the simulations

```{r}
#| label: raster into polygons
lc_poly_1990 <- as.polygons(rc_lc_1990)

# turning it into an sf object
lc_poly_sf <- st_as_sf(lc_poly_1990)

#dissolving the polygons so we can do an area calculation
lc_poly_sf <- lc_poly_sf |>
  group_by(U2000_CLC1990_V2020_20u1) |>
  summarise()

lc_poly_sf <- st_transform(lc_poly_sf, 2157)
lc_poly_sf <- st_intersection(lc_poly_sf, ireland) #clipping the polygons so the borders are the same as my viewing window

mapview(lc_poly_sf, zcol = "U2000_CLC1990_V2020_20u1")

```



Generating test statistic!
```{r}
#| label: barrow counts by habitat
#making lc into a tesselation
lc_list <- split(lc_poly_sf, lc_poly_sf$U2000_CLC1990_V2020_20u1) #grabbing individual polygons
lc_tiles <- lapply(lc_list, as.owin) #each one as a window

#tesselation of the polygon windows
lc_tess <- tess(tiles = lc_tiles)

#now the test statistic
obs_barrow_counts <- quadratcount(barrows_ppp, tess=lc_tess) |> as.vector()
names(obs_barrow_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
obs_barrow_counts <- obs_barrow_counts |> as_tibble_row()
```

The test statistic: Referencing the literature, it is believed that fairies prefer forested habitats, which offer readily available hollow trees, babbling streams, and mossy grottos. Our first hypothesis, therefore, is that fairies prefer to make their homes in forests.

```{r}
obs_barrow_counts
```


habitat 'intensity', which gives the proportional probability of any point existing in each habitat type (based on area)

#| label: habitat intensity function
tile_areas <- sapply(lc_tiles, area.owin) #the area of each tile
tile_probs <- tile_areas / sum(tile_areas) #getting the proportions

habitat_prob_intensity <- function(x, y) { #intensity function for probability of points being in each habitat
  sapply(seq_along(lc_tiles), function(i) {
    inside.owin(x, y, lc_tiles[[i]]) * tile_probs[i]
  }) |> rowSums()
}


```{r}
#new habitat prob intensity, to try and make this run faster
tile_areas <- sapply(lc_tiles, area.owin) #the area of each tile
tile_probs <- tile_areas / sum(tile_areas) #getting the proportions

habitat_prob_intensity <- function(x, y) {
  coords <- cbind(x, y)
  cell_vals <- terra::extract(rc_lc_1990, coords)[,1] #pulls out the habitat value the point lands on top of
  cell_vals[is.na(cell_vals) | is.nan(cell_vals) | cell_vals < 1 | cell_vals > length(tile_probs)] <- 1 #prevent error by keeping everything in range

  probs <- tile_probs[cell_vals] #assigning the probabilties
  
  probs[is.na(probs)] <- 0 #preventing errors with NA by making them 0
  return(probs)
}
```

## Monte Carlo Simulations: Barrows

```{r}

num_barrows <- nrow(barrows_sf)#to set number of points to sample

#creating a function that will plot barrows around ireland according to the area proportions of the various habitats
#and checking the runtime to make sure we can do the full 999 sims
runtime <- system.time({gen_barrows_sim <- function() {
  barrows_sim <- rpoint(n=num_barrows,f=habitat_prob_intensity, win=barrows_ppp$window)
  return(barrows_sim) 
}
single_barrows_sim_ppp <- gen_barrows_sim()
#IT WORKED!!! YAYYYYYY
single_barrow_sim_counts <- quadratcount(single_barrows_sim_ppp, tess=lc_tess) |> as.vector()
names(single_barrow_sim_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
single_barrow_sim_counts <- single_barrow_sim_counts |> as_tibble_row()})
single_barrow_sim_counts
runtime
```



This is what I ran to do the monte carlo simulations. I never want to run that again, but here is the code
OK and I do have to run this again and save sims_list so that we have the points for the distance calculation, unfortunately

```{r}
library('pbapply')
#Now we are running 999 simulations wahoooo

compute_quadrat_counts <- function(sim_ppp) { #making a function to apply to all the sims
  sim_counts <- quadratcount(sim_ppp, tess=lc_tess) |> as.vector()
  names(sim_counts) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies")
  return(sim_counts)
}

gen_sims_ppp <- function(num_sims) { #function to generate 999 ppp simulations
  barrows_sims <- spatstat.random::rpoint(
    n = num_barrows,
    f = habitat_prob_intensity,
    win = barrows_ppp$window,
    nsim = num_sims
  )
  return(barrows_sims)
}

sims_list <- gen_sims_ppp(999) #simulations
sims_habitat_counts <- pbapply::pblapply(sims_list, FUN=compute_quadrat_counts) #apply counts function, with a progress bar from pbapply
full_count_df <- as_tibble(sims_habitat_counts) |> t() |> as_tibble() #smoosh them together
colnames(full_count_df) <- c("Dense Human Habitat", "Human Agriculture", "Forests", "Grass and Shrublands", "Rocky, Barren, or Snowy", "Wetlands", "Water Bodies") #label columns
full_count_df |> head() #tada!

#this took two hours
```


```{r}
#write_csv(full_count_df, "data/raw/monte_carlo_sims.csv") #done after code above
#saveRDS(sims_list, "data/raw/sims_ppplist.rds")

#reading them back in after saving
monte_carlo_df <- read_csv('data/raw/monte_carlo_sims.csv')
barrows_sims_ppplist <- readRDS("data/raw/sims_ppplist.rds")

```

# Hypothesis Testing

## Hypothesis 1: Barrow Preferred Habitats

Alright now we need to test hypothesis 1 against the test statistic from earlier. Hypothesis 1 is something like 'fairies prefer forest habitat' because of a couple poems I'll write down later. I have a sneaking suspicion that this will not end up being true, so I will likely run the hypothesis test on every habitat type

Let's plot those statistics, shall we

```{r}
#have to rename my categories actually so they work with the histograms
names(monte_carlo_df) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")

names(obs_barrow_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
```

```{r}
#| label: forests distribution
  mc_df <- bind_rows(monte_carlo_df, obs_barrow_counts)
  monte_carlo_df |> ggplot(aes(x=Forests)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Forests, color = "darkgreen") +
    theme_classic()
#hahahaha turns out that's way lower, so they actually AVOID forests. What a fun finding!

```

```{r}
#| label: dense human distribution
  monte_carlo_df |> ggplot(aes(x=Dense_Human_Habitat)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Dense_Human_Habitat, color = "red") +
    theme_classic()
#We're pretty close to the bounds for that one

```


```{r}
#| label: human agriculture
  monte_carlo_df |> ggplot(aes(x=Human_Agriculture)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Human_Agriculture, color = "orange") +
    theme_classic()
#WOW that's way higher, which is not what I was expecting. Will have to comment on this

```

```{r}
#| label: grass and shrublands
  monte_carlo_df |> ggplot(aes(x=Grass_and_Shrublands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Grass_and_Shrublands, color = "green") +
    theme_classic()
#We're pretty close to the bounds for that one too

```

```{r}
#| label: rocky, barren, and snowy
  monte_carlo_df |> ggplot(aes(x=Rocky_Barren_Snowy)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Rocky_Barren_Snowy) +
    theme_classic()
#Far below, they DONT like these spots

```

```{r}
#| label: wetlands
  monte_carlo_df |> ggplot(aes(x=Wetlands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Wetlands, color = "lightblue") +
    theme_classic()
#Far below, they DONT like these spots either

```

```{r}
#| label: water bodies 2
  monte_carlo_df |> ggplot(aes(x=Water_Bodies)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_barrow_counts$Water_Bodies, color = "blue") +
    theme_classic()
#Far below, they DONT like these spots either either

```

It appears that they might have clustering, in human agriculture, and avoid some of the other areas! (of course this technically makes sense because 'fairy barrows' are actually ancient human burial mounds, so they would be found in populated areas, but we're talking about fairies here)

# P value calculations
```{r}
#| label: p value calculations

human_hab_count_more_extreme <- mc_df |> filter(Dense_Human_Habitat >= obs_barrow_counts$Dense_Human_Habitat)
human_hab_p_val <- nrow(human_hab_count_more_extreme) / nrow(mc_df)

ag_count_more_extreme <- mc_df |> filter(Human_Agriculture >= obs_barrow_counts$Human_Agriculture)
ag_p_val <- nrow(ag_count_more_extreme) / nrow(mc_df)

forest_count_more_extreme <- mc_df |> filter(Forests >= obs_barrow_counts$Forests)
forest_p_val <- nrow(forest_count_more_extreme) / nrow(mc_df)

grass_shrub_count_more_extreme <- mc_df |> filter(Grass_and_Shrublands >= obs_barrow_counts$Grass_and_Shrublands)
grass_shrub_p_val <- nrow(grass_shrub_count_more_extreme) / nrow(mc_df)

barren_count_more_extreme <- mc_df |> filter(Rocky_Barren_Snowy >= obs_barrow_counts$Rocky_Barren_Snowy)
barren_p_val <- nrow(barren_count_more_extreme) / nrow(mc_df)

wetland_count_more_extreme <- mc_df |> filter(Wetlands >= obs_barrow_counts$Wetlands)
wetland_p_val <- nrow(wetland_count_more_extreme) / nrow(mc_df)

water_count_more_extreme <- mc_df |> filter(Water_Bodies >= obs_barrow_counts$Water_Bodies)
water_p_val <- nrow(water_count_more_extreme) / nrow(mc_df)



#I could make this into a df instead of doing this way huh?

print("P-values by habitat:")
print(paste("Dense Human Habitat:",human_hab_p_val))
print(paste("Human Agriculture:",ag_p_val))
print(paste("Forest:",forest_p_val))
print(paste("Grass and Shrublands:",grass_shrub_p_val))
print(paste("Rocky, Barren, and Snowy:",barren_p_val))
print(paste("Wetlands:",wetland_p_val))
print(paste("Water Bodies:",water_p_val))

#so out of all of these only agriculture is truly significant, but all the '1'basically means the true count was WAY lower than the expected count, which is also significant, and I'll work on that in a bit
p_values <- c(human_hab_p_val, ag_p_val, forest_p_val, grass_shrub_p_val, barren_p_val, wetland_p_val,water_p_val)
p_val_greater_df <- data.frame(habitat = lc_labels, p_value = p_values)
print(p_val_greater_df)

```

Something something despite stories, perhaps fairy hunting grounds are different than their preferred living situations, maybe should have another hypothesis that tests whether they avoid certain habitats for their homes (which would be forests, barren, wetlands, water bodies dont really count)

So! Onto hypothesis 2: Do they have any relationship to the distance from stone circles aka portals to the fairy realm?

I think we need to get the mean distance of each point to its CLOSEST stone circle, which sounds like something I should be able to do

Yes, we can do nearest neighbor calculations and select only the minimum distance ones

ah gosh dangit, I never saved the simulated points, only the counts. That's awkward
saved them!

## Hypothesis 2: Proximity to Stone Circles

Now we are going to have to run some distence calculations to see if barrows cluster around stone circles. 
```{r}
#| label: calculating test statistic of nearest neighbor distances
obs_dists_sc <- nncross(barrows_ppp, stone_circles_ppp)$dist
mean_obs_dists <- mean(obs_dists_sc)
```

```{r}
mean_obs_dists
```

```{r}
#| label: creating the distance function for the whole dataset
compute_mean_dist <- function(sim_ppp) {
  sim_dists <- nncross(sim_ppp, stone_circles_ppp)$dist
  mean(sim_dists)
}

sim_distances <- lapply(X=barrows_sims_ppplist, FUN=compute_mean_dist)

```

```{r}
#| label: visualizing mean distance null distribution

sim_dist_df <- unlist(sim_distances) |> as_tibble()
obs_dist_df <- tibble(value = mean_obs_dists)
all_dists_df <- rbind(sim_dist_df,obs_dist_df)
all_dists_df |> ggplot(aes(x=value)) +
  geom_histogram(fill='black', alpha=0.5) +
  geom_vline(xintercept = mean_obs_dists, color='darkorchid4') +
  theme_classic()
```

Interpretation: The fairies appear to live much closer to the stone circles than predicted

```{r}
#| label: computing distance p value

count_more_extreme <- all_dists_df |> filter(value >= mean_obs_dists)
sc_dist_p_val <- nrow(count_more_extreme) / nrow(all_dists_df)
sc_dist_p_val
```

basically all of the dataset is more extreme than the observed distance statistic. Wait, this would mean that they are FAR closer to the stone circles than they would be if randomly distributed. WOOOOO!

## Hypothesis 3: Stone Circle Primary Habitats
Now that we have determined that stone circle location is related to where fairies create their homes, and with resounding theories that they form in habitats that are similar to that of the fairy realm, let us run a simulation to see if the stone circles appear in any habitat more than the rest. We must preserve these locations

## Monte Carlo Simulations: Stone Circles

```{r}
num_circles <- nrow(stone_circles_sf) #to set number of points to sample

#creating a function that will plot barrows around ireland according to the area proportions of the various habitats
#and checking the runtime to make sure we can do the full 999 sims
runtime <- system.time({gen_circles_sim <- function() {
  circles_sim <- rpoint(n=num_circles,f=habitat_prob_intensity, win=stone_circles_ppp$window)
  return(circles_sim) 
}
single_circles_sim_ppp <- gen_circles_sim()
#IT WORKED!!! YAYYYYYY
single_circle_sim_counts <- quadratcount(single_circles_sim_ppp, tess=lc_tess) |> as.vector()
names(single_circle_sim_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
single_circle_sim_counts <- single_circle_sim_counts |> as_tibble_row()})
single_circle_sim_counts
runtime

```

now to take a little while again

```{r}
library('pbapply')
#Now we are running 999 simulations wahoooo

compute_quadrat_counts <- function(sim_ppp) { #making a function to apply to all the sims
  sim_counts <- quadratcount(sim_ppp, tess=lc_tess) |> as.vector()
  names(sim_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
  return(sim_counts)
}

gen_sims_ppp <- function(num_sims) { #function to generate 999 ppp simulations
  stone_circle_sims <- spatstat.random::rpoint(
    n = num_circles,
    f = habitat_prob_intensity,
    win = stone_circles_ppp$window,
    nsim = num_sims
  )
  return(stone_circle_sims)
}

sc_sims_list <- gen_sims_ppp(999) #simulations
sc_habitat_counts <- pbapply::pblapply(sc_sims_list, FUN=compute_quadrat_counts) #apply counts function, with a progress bar from pbapply
sc_count_df <- as_tibble(sc_habitat_counts) |> t() |> as_tibble() #smoosh them together
colnames(sc_count_df) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies") #label columns
sc_count_df |> head() #tada!

#this took two hours
```

```{r}
#saving so we dont have to do that again either
#write_csv(sc_count_df, "data/raw/sc_monte_carlo_sims.csv") #done after code above
sc_monte_carlo_df <- read_csv('data/raw/sc_monte_carlo_sims.csv')
```

Now to generate that test statistic, we are slightly out of order here

```{r}
obs_sc_counts <- quadratcount(stone_circles_ppp, tess=lc_tess) |> as.vector()
names(obs_sc_counts) <- c("Dense_Human_Habitat", "Human_Agriculture", "Forests", "Grass_and_Shrublands","Rocky_Barren_Snowy","Wetlands","Water_Bodies")
obs_sc_counts <- obs_sc_counts |> as_tibble_row()
```

```{r}
obs_sc_counts
```

And the distributions for all habitats

```{r}
#| label: forests distribution stone circles
  sc_df <- bind_rows(sc_count_df, obs_sc_counts)
  sc_count_df |> ggplot(aes(x=Forests)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Forests, color = "darkgreen") +
    theme_classic()
#right on target, so nothing special there

```

```{r}
#| label: dense human distribution stone circles
  sc_count_df |> ggplot(aes(x=Dense_Human_Habitat)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Dense_Human_Habitat, color = "red") +
    theme_classic()
#also just about right

```


```{r}
#| label: human agriculture stone circles
  sc_count_df |> ggplot(aes(x=Human_Agriculture)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Human_Agriculture, color = "orange") +
    theme_classic()
#right in the middle

```

```{r}
#| label: grass and shrublands stone circles
  sc_count_df |> ggplot(aes(x=Grass_and_Shrublands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Grass_and_Shrublands, color = "green") +
    theme_classic()
#high for this one! nice!

```

```{r}
#| label: rocky, barren, and snowy stone circles
  sc_count_df |> ggplot(aes(x=Rocky_Barren_Snowy)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Rocky_Barren_Snowy) +
    theme_classic()
#eh maybe higher, not a great distribution though

```

```{r}
#| label: wetlands stone circles
  sc_count_df |> ggplot(aes(x=Wetlands)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Wetlands, color = "lightblue") +
    theme_classic()
#in the middle

```

```{r}
#| label: water bodies
  sc_count_df |> ggplot(aes(x=Water_Bodies)) +
    geom_bar(stat='count') +
    geom_vline(xintercept = obs_sc_counts$Water_Bodies, color = "blue") +
    theme_classic()
#Far below, makes sense though, maybe put something about avoiding running water

```

It appears that there may be clustering in grasslands! Or at least a slight preference

# P value calculations for stone circles
```{r}
#| label: p value calculations for stone circle

human_hab_count_more_extreme_sc <- sc_df |> filter(Dense_Human_Habitat >= obs_sc_counts$Dense_Human_Habitat)
human_hab_p_val_sc <- nrow(human_hab_count_more_extreme_sc) / nrow(sc_df)

ag_count_more_extreme_sc <- sc_df |> filter(Human_Agriculture >= obs_sc_counts$Human_Agriculture)
ag_p_val_sc <- nrow(ag_count_more_extreme_sc) / nrow(sc_df)

forest_count_more_extreme_sc <- sc_df |> filter(Forests >= obs_sc_counts$Forests)
forest_p_val_sc <- nrow(forest_count_more_extreme_sc) / nrow(sc_df)

grass_shrub_count_more_extreme_sc <- sc_df |> filter(Grass_and_Shrublands >= obs_sc_counts$Grass_and_Shrublands)
grass_shrub_p_val_sc <- nrow(grass_shrub_count_more_extreme_sc) / nrow(sc_df)

barren_count_more_extreme_sc <- sc_df |> filter(Rocky_Barren_Snowy >= obs_sc_counts$Rocky_Barren_Snowy)
barren_p_val_sc <- nrow(barren_count_more_extreme_sc) / nrow(sc_df)

wetland_count_more_extreme_sc <- sc_df |> filter(Wetlands >= obs_sc_counts$Wetlands)
wetland_p_val_sc <- nrow(wetland_count_more_extreme_sc) / nrow(sc_df)

water_count_more_extreme_sc <- sc_df |> filter(Water_Bodies >= obs_sc_counts$Water_Bodies)
water_p_val_sc <- nrow(water_count_more_extreme_sc) / nrow(sc_df)


p_values_sc <- c(human_hab_p_val_sc, ag_p_val_sc, forest_p_val_sc, grass_shrub_p_val_sc, barren_p_val_sc, wetland_p_val_sc,water_p_val_sc)
p_val_greater_df_sc <- data.frame(habitat = lc_labels, p_value = p_values_sc)
print(p_val_greater_df_sc)

```

Excellent, so this provides evidence that fairies prefer to build their circles in grass and shrublands!

# Land Cover Change Analysis

Time to see if we have lost grass and shrubland area! Thankfully we already have the 1990 reclassified raster, and need to do the same thing for 2018

```{r}
#| label: land cover change analysis, grasslands
#reclassifying the 2018 data
rc_lc_2018 <- classify(lc_2018, rclmat, include.lowest=TRUE)
plot(rc_lc_1990, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 1990")
plot(rc_lc_2018, col = lc_colors, type = "classes", main = "Reclassified Land Cover: 2018")

#realigning the pixels to make sure they're on top of each other
rc_lc_2018 <- resample(rc_lc_2018, rc_lc_1990, method = "near")


#stacking the raster so the pixels are paired
lc_stack <- c(rc_lc_1990, rc_lc_2018)
names(lc_stack) <- c("1990", "2018")

#changing the column/row names
lc_class_names <- c(
  "1" = "Dense_Human_Habitat",
  "2" = "Human_Agriculture",
  "3" = "Forests",
  "4" = "Grass_and_Shrublands",
  "5" = "Rocky_Barren_Snowy",
  "6" = "Wetlands",
  "7" = "Water_Bodies"
)

change_matrix <- crosstab(lc_stack, long = FALSE)
colnames(change_matrix) <- lc_class_names[colnames(change_matrix)]
rownames(change_matrix) <- lc_class_names[rownames(change_matrix)]

change_matrix

```

Now that we have the change matrix (which I'll make a pretty chart for), I want to map the pixels that started as grasslands, became grasslands, or grasslands that were lost

```{r}
# we know our grass and shrublands are class 4
grass_code <- 4

#this got complicated quick, but basically I had to make an ireland-shaped polygon of nothing, with pixels where grassland exists
grass_1990_r <- rc_lc_1990
values(grass_1990_r) <- ifelse(is.na(values(rc_lc_1990)), NA,
                               values(rc_lc_1990) == grass_code)

grass_2018_r <- rc_lc_2018
values(grass_2018_r) <- ifelse(is.na(values(rc_lc_2018)), NA,
                               values(rc_lc_2018) == grass_code)

# now I have to make a little transition raster that's also an empty ireland-shaped polygon of nothing
transition <- rc_lc_1990
values(transition) <- NA

# and make sure we only update the cells that are inside the original land cover raster (this got so technical)
valid <- !is.na(values(rc_lc_1990))

#now I'm making my new transition classes all about grassland
values(transition)[valid & values(grass_1990_r) & values(grass_2018_r)] <- 1  #stayed as grassland
values(transition)[valid & !values(grass_1990_r) & values(grass_2018_r)] <- 2 #new grassland
values(transition)[valid & values(grass_1990_r) & !values(grass_2018_r)] <- 3 #lost grassland
values(transition)[valid & !values(grass_1990_r) & !values(grass_2018_r)] <- 0 #never grassland

#and plotting
transition_colors <- c("grey80","darkgreen","lightgreen","saddlebrown")
plot(transition, col = transition_colors, type="classes",
     main="Grass and Shrublands Transitions", axes=TRUE)

```

And I want the real numbers of the transition, so we're going back to the change matrix to grab those

```{r}
persistent <- change_matrix["Grass_and_Shrublands","Grass_and_Shrublands"]
gain <- sum(change_matrix[-which(rownames(change_matrix)=="Grass_and_Shrublands"),
                         "Grass_and_Shrublands"]) #anything that didnt have grass in 1990 but does in 2018

loss <- sum(change_matrix["Grass_and_Shrublands",
                          -which(colnames(change_matrix)=="Grass_and_Shrublands")]) #grass in 1990 but not in 2018

never_grass <- sum(change_matrix[-which(rownames(change_matrix)=="Grass_and_Shrublands"),
                               -which(colnames(change_matrix)=="Grass_and_Shrublands")]) #never grass

grass_summary <- c(Persistent = persistent,
                   Gain = gain,
                   Loss = loss,
                   Never_Grass = never_grass)
print(grass_summary)

#and checking that the raster has the proper pixels, which it does!
pixel_counts <- freq(transition)

print(pixel_counts)


```

in the 1990 land cover data, 153237 of the 20780550 pixels were grass and shrubland. This is 0.737% of the total pixels.
172092 pixels were grass and shrubland in 2018, which is 0.828% of the total pixels. We're dealing with very small numbers here, and a very limited habitat for stone circle construction!